---
date: "July 21, 2022"
title: "Employment data from BLS, ESD, PSRC"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = here::here("pptx")) })
output: 
  officedown::rpptx_document: 
    reference_doc: "../pptx/OERF_template_20220721_19pt.pptx"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE)
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

# fig_width_smaller  <- 6.25
# fig_height_smaller <- 4.35
# fig_pos_x_smaller  <- 6.6
# fig_pos_y_smaller  <- 1.7

fig_w_stack <- 6.8
fig_h_stack <- 5
fig_x_stack <- 6
fig_y_stack <- 1.6

fig_width_smaller  <- 6.75
fig_height_smaller <- 4.5
fig_pos_x_smaller  <- 6.0
fig_pos_y_smaller  <- 1.7

fig_width_bigger  <- 7.5
fig_height_bigger <- 5.0
fig_pos_x_bigger  <- 2.9
fig_pos_y_bigger  <- 1.7

fig_width_wider  <- 12.5
fig_height_wider <- 5.0
fig_pos_x_wider  <- 0.42
fig_pos_y_wider  <- 1.6

source(here::here("scripts", "00-initialize.R"), local = knitr::knit_global())

theme_set(theme_slides_bigger)

path_ihs_main <- "//ad.seattle.gov/dept/ERF/Data/OERF/Forecast/Economic/IHS/"
path_cbo_ksm  <- "//ad.seattle.gov/dept/ERF/Data/OERF/Forecast/Economic/CBO/Models/KSM/"
path_oerf_ksm <- "//ad.seattle.gov/dept/ERF/Data/OERF/Forecast/Economic/OERF/Models/KSM/"
```

```{r palettes}
# pal <- wes_palette("GrandBudapest1")[c(2, 3)]
# pal <- pnw_palette("Bay", 4, type ="continuous")[c(1, 4)]
# pal <- viridis_pal(option = "D", begin = 0.65, end = 0.25)(2)[c(1, 2)]

pal_ks_vs_wa_vs_us_blue_green_red <- pnw_palette("Bay", 4, type ="continuous")[c(1, 2, 4)]
pal_ks_vs_wa_base <- c("blue", "darkgreen", "red")

pal_ks_vs_us_blue_red <- pnw_palette("Bay", 4, type ="continuous")[c(4, 1)]

pal_ks_vs_us_blue_green <- viridis_pal(option = "D", begin = 0.65, end = 0.25)(2)[c(1, 2)]

pal_ks_vs_us <- pal_ks_vs_us_blue_red

# pal <- brewer.pal(9, "Set1")[c(1, 2)]
# pal <- pnw_palette("Sunset", 3, type ="continuous")[c(2, 1)]
# pal <- pnw_palette("Sunset2", 3, type ="continuous")[c(2, 1)]
# pal <- pnw_palette("Bay", 5, type ="continuous")[c(5, 1)]

pal_scenarios_3_ylorrd <- brewer.pal(9, "YlOrRd")[c(9, 6, 3)]
pal_scenarios_3_mrfox <- wes_palette("FantasticFox1")[c(5, 4, 3)]
pal_scenarios_3_budapest <- wes_palette("GrandBudapest1")[c(2, 1, 3)]
pal_scenarios_2_pink_blue <- c(wes_palette("GrandBudapest1")[2], pnw_palette("Bay", 1, type ="continuous"))
pal_scenarios_2_lightblue_blue <- c("#CCDDEA", "#2F5597")
                                 
# pal_scenarios <- pal_scenarios_3_ylorrd
pal_scenarios <- pal_scenarios_3_budapest

pal_2019_vs_2021_yellow_orange <- pnw_palette("Bay", 4, type ="continuous")[c(3, 4)]
pal_2019_vs_2021_green_blue <- c("#A9D18E", "#2F5597")

pal_2019_vs_2021 <- pal_2019_vs_2021_green_blue
```



## U.S. BLS employment data: CES, CPS, LAUS

**Current Population Survey (CPS)**

- Also referred to as household survey
- Monthly survey of roughly 60,000 households with approximately 110,000 individuals
- Provides data on the employment, unemployment, labor force, persons not in the labor force, hours of work, earnings, and other characteristics

**Current Employment Statistics (CES)**

- Also referred to as payroll or establishment survey
- Monthly survey of approximately 131,000 businesses and government agencies, representing 670,000 individual worksites
- Provides industry estimates of employment, hours, and earnings of workers on nonfarm payrolls

<!--
- Data are gathered based on employment in the Sunday-to-Saturday week that includes the 12th of the month
-->

<br>

- https://www.wsj.com/articles/diverging-jobs-data-raise-questions-about-labor-market-health-11658136600



## U.S. BLS employment data: CES, CPS, LAUS

**Local Area Unemployment Statistics (LAUS)**

- Data from several sources is used to produce monthly employment, unemployment, and labor force data for states, counties, metropolitan areas, and many cities, by place of residence
- Concepts and definitions underlying LAUS data come from the CPS household survey

<br>

- State monthly model-based estimates are controlled to sum to national monthly employment and unemployment estimates from the CPS. These models combine current and historical data from the CPS and CES surveys, and from state unemployment insurance (UI) systems.
- County estimates are produced using data from several sources, including CPS, CES, state UI systems, and the Census Bureau's American Community Survey (ACS), to create estimates that are adjusted to the statewide measures of employment and unemployment
- Estimates for cities are prepared from county data using dis-aggregation techniques based on inputs from the ACS, population estimates, employment, and UI claims counts



## U.S. BLS employment data CES, CPS, LAUS

```{r ph=officer::ph_location(width = fig_width_wider, height = fig_height_wider, left = fig_pos_x_wider, top = fig_pos_y_wider)}
bls_emp_plt <- 
  dir_ls(here("data", "wrangled"), regex = "bls_emp_plt") %>% 
  last() %>% 
  read_csv() %>% 
  mutate(yearm = as.yearmon(yearm))

g_bls_emp <- 
  ggplot() +
    aes(x = yearm, y = value, col = data_survey_type_label, # alpha = seas, linetype = seas, 
               text = hover_label, group = var_code) +
    geom_line(size = 1) +
    scale_y_continuous(labels = label_number(big.mark = ",")) +
    scale_alpha_manual(values = c(1, 0.5)) +
    labs(x = NULL, y = NULL, col = NULL, alpha = NULL, linetype = NULL,
         caption = "Source: U.S. Bureau of Labor Statistics")  +
  facet_wrap(~ area_label, ncol = 1, scales = "free_y")+
  theme(legend.margin = margin(0, 0, -10, 0, unit = "pt")) 
  
g_bls_emp_lvl <- 
  g_bls_emp %+% 
  {bls_emp_plt %>% 
      filter(seas == "NSA",
             measure == "lvl")} +
  labs(title = "Employment, Not seasonally adjusted")

dml(ggobj = g_bls_emp_lvl)
```


  
  
## U.S. BLS employment data CES, CPS, LAUS

```{r ph=officer::ph_location(width = fig_width_wider, height = fig_height_wider, left = fig_pos_x_wider, top = fig_pos_y_wider)}
g_bls_emp_chng <- 
  g_bls_emp %+% 
  {bls_emp_plt %>% 
  filter(seas == "NSA",
         measure == "dmm",
         yearm >= "Jan 2021")} +
  geom_hline(yintercept = 0, col = "gray50") +
  labs(title = "Change in employment from previous month, Not seasonally adjusted")

dml(ggobj = g_bls_emp_chng)
```




## WA ESD employment data: CES, LAUS, QCEW, WA-QB

**Washington employment estimates (CES)**

- CES statistics are developed by BLS from a monthly survey of about 7,000 businesses and government agencies in Washington, which cover around 20,000 individual worksites 
- See BLS CES slide above for more detail
- Data: https://esd.wa.gov/labormarketinfo/employment-estimates

<br>


**Local Area Unemployment Statistics (LAUS)**

- See BLS LAUS slide above for more detail
- Data: https://esd.wa.gov/labormarketinfo/labor-force



## WA ESD employment data: CES, LAUS, QCEW, WA-QB

**Quarterly Census of Employment and Wages (QCEW)**

- Also referred to as covered employment data
- Administrative records from quarterly unemployment tax forms filed by employers to the Washington State Employment Security Department (ESD) for unemployment insurance reporting under Washington Unemployment Insurance Act
- In addition, ESD conducts quarterly surveys of employers who have more than one worksite in the state
- ESD uses these records and surveys to publish about six moths after the end of the quarter a dataset that includes monthly employment and quarterly wages paid, by county and industry
- Unit of measurement is jobs, rather than working persons, part-time and temporary positions are included, if a worker holds more than one job, each job would appear separately
- Jobs exempt from unemployment insurance: unincorporated self-employed, jobs paid on commission, CEOs and corporate officers, uniformed military, elected officials, church jobs, railroad workers
- Typically, covered employment has represented about 85-90% of total employment
- Data: https://esd.wa.gov/labormarketinfo/covered-employment



## WA ESD employment data: CES, LAUS, QCEW, WA-QB
 
**Washington employment estimates (WA-QB)**

- Monthly estimates of non-farm employment, by industry and metro area (not-seasonally adjusted also available for major counties)
- Employment estimates are based on CES data from BLS, but are different from those in the CES data series
- To develop employment estimates ESD replaces CES survey data with data from the QCEW as follows
  - Each month, economists estimate monthly job gains and losses based on CES
  - At the end of each quarter, economists revise the estimates based on actual numbers from employer tax records in QCEW
  - Recall: QCEW data are actual counts of employees, CES eestimates come from a survey of employers
  - Replacing employment estimates with the actual number of job gains or losses is called benchmarking
  - ESD benchmarks data quarterly, the BLS benchmarks its data only once a year
- Data: https://esd.wa.gov/labormarketinfo/employment-estimates



## ESD WA-QB employment data revisions after benchmarking

```{r ph=officer::ph_location(width = fig_width_wider, height = fig_height_wider, left = fig_pos_x_wider, top = fig_pos_y_wider)}
esd_qb_combined_plt <- 
  read_csv(here("data", "wrangled", "ks_esd_qb_combined_plt_202206.csv")) %>% 
  mutate(yearm = as.yearmon(yearm),
         release = as_factor(release))

n_colors <- 8

pal_esd_benchmarks_pmw <- pnw_palette("Bay", n_colors) %>% rev()
pal_esd_benchmarks_base <- c("darkgreen", "blue", "cyan3", "orange", "red", "purple", "brown", "yellow")

pal_esd <- pal_esd_benchmarks_base

plot_esd_emp <- function(.var_code, .label) {
  esd_qb_selected_plt <- 
    esd_qb_combined_plt %>%
    filter(var_code %in% .var_code,
           yearm >= "Jan 2019") 
  
  area_label <- 
    case_when(str_sub(.var_code, 1, 2) == "KS" ~ "Seattle MD",
              str_sub(.var_code, 1, 2) == "WA" ~ "Washington State")

  g_esd <- 
    esd_qb_selected_plt %>%
    ggplot() +
      aes(x = yearm, y = value, col = release, alpha = release, text = hover_label, group = release) +
      geom_line(size = 1) +
      scale_color_manual(values = rep(pal_esd, each = 3)) +
      scale_alpha_manual(values = rep(c(0.4, 0.7, 1.0), times = n_colors)) +
      scale_y_continuous(labels = label_number(big.mark = ",")) +
      labs(x = NULL, y = NULL, col = NULL, alpha = NULL,
           title = str_c(area_label, " Employment by ESD Release - ", .label)) +
      guides(color = guide_legend(nrow = 3, byrow = FALSE))
  
  g_esd + facet_wrap(~seas_label, ncol = 2)
}

g_esd_emp_ttl <- 
  plot_esd_emp("KS_N", "Total Nonfarm")
g_esd_emp_inf <- 
  plot_esd_emp("KS_NINF", "Information")

dml(ggobj = g_esd_emp_ttl)
```



## PSRC covered employment data based on ESD QCEW

- Puget Sound Regional Council (PSRC) covered employment by city is constructed using QCEW data from ESD
- Data is released in July for March of the previous year
- March is used as a representative month when seasonal fluctuations are minimized
- PSRC assigns geographic location to the records based on site address
- PSRC supplements ESD data using a survey of public sector employers, Boeing, Microsoft, to disaggregate headquarters employment to actual multiple workplaces
- Estimates are proportionally adjusted to match ESD estimates for March at the county and major sector level, to compensate for records that were removed from the ESD data because they could not be precisely sited
- Private sector jobs are categorized by NAICS; public sector jobs are categorized separately as “Education” (public schools and colleges) or “Government” (all else)
- Where aggregate employment values represent fewer than three reporting firms, or when a single employer accounts for more than 80% of jobs, the value is withheld
- Data: https://psrc.org/covered-employment-estimates



## PSRC covered employment data based on ESD QCEW

```{r ph=officer::ph_location(width = fig_width_wider, height = fig_height_wider, left = fig_pos_x_wider, top = fig_pos_y_wider)}
ks_qcew_naics_2_plt <- 
  read_csv(here("data", "wrangled", "ks_qcew_naics_2_plt.csv")) %>% 
  mutate(across(c(area_label, naics_custom, naics_custom_label), as_factor),
         area_label = fct_rev(area_label))

g_psrc_facet_by_naics <- 
 
g_psrc_trd_and_inf <- 
  ks_qcew_naics_2_plt %>%
  filter(data_source == "PSRC",
         area_label %in% c("Seattle MD", "Bellevue", "Seattle"),
         measure == "Mar Employment",
         naics_custom %in% c("42, 44-45", "51")) %>% 
  filter(yr >= 2011) %>% 
  ggplot() +
    aes(x = yr, y  = gyy, col = area_label) +
    geom_hline(yintercept = 0, col = "gray50") +
    geom_line(size = 1) +
    scale_x_continuous(breaks = 2011:2021) +
    scale_y_continuous(limits = c(-0.1, 0.3),
                       labels = label_percent()) +
    scale_linetype_manual(values = c(2, 1)) +
    labs(x = NULL, y = NULL, col = NULL, 
         title = "Employment growth for trade and information sectors, March 2011 to March 2021") +
    facet_wrap(~naics_custom_label, scales = "free_x")

dml(ggobj = g_psrc_trd_and_inf)
```

